---
title: "DNA methylation analysis of SMC phase"
author: "Kennedy Zinn"
format: html
editor: visual
---

```{r, echo = FALSE}
suppressPackageStartupMessages(library(minfi))
suppressPackageStartupMessages(library(limma))
suppressPackageStartupMessages(library(DMRcate))
suppressPackageStartupMessages(library(maxprobes))
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(missMethyl))
suppressPackageStartupMessages(library(MethylToSNP))
```

# Smooth Muscle Cell (SMC) Comparisons

## Load Data

load the idat files: For SMC comparisons, samples 2 APCK, 4 APCK, 6 APCK, 8 APCK, 10 APCK, 12 APCK and 14 APCK are from SMC phase (rest are from neural crest phase). Of those, 2, 6, 10 and 14 are the controls.

```{r, echo = FALSE}
rgSet <- read.metharray.exp("/Users/kennedyzinn/Desktop/OneDrive - UTHealth Houston/GRA/dna methyl/SMC", force = T)
```

load sample sheet

```{r, echo = FALSE}
sample_sheet <- read_excel("/Users/kennedyzinn/Desktop/OneDrive - UTHealth Houston/GRA/dna methyl/SMC/SampleSheet_SMC.xlsx")
```

## Quality check

```{r, echo = FALSE}
suppressMessages(qcReport(rgSet, pdf = "/Users/kennedyzinn/Downloads/qcReport.pdf", sampNames = sample_sheet$`Sample Name`, sampGroups = sample_sheet$Status))
```

## Pre-Process
```{r}
source("/Users/kennedyzinn/Desktop/OneDrive - UTHealth Houston/GRA/dna methyl/process_methyl_microarray.R")
gmSet <- process_methyl_microarray(rgSet)
```

Beta values, which range between [0,1], represent the proportion of methylated DNA copies at a specific CpG site compared to all of the DNA copies at that site.  Beta values for the ith CpG site are mathematically defined as,

Beta(i) = max(yi(methyl),0) / [max(yi(unmethyl), 0) + max(yi(methyl), 0) + alpha],

where yi methyl and unmethyl are the intensities measured by the ith methylated and unmethylated probes, respectively. 0 ensures that negative signals that take place after background adjustment are not considered. Alpha, a constant (usually 100), serves as an offset to regularize beta when both methylated and unmethylated instensities are low. Beta values are intuitive, and hence useful for visual presentations of the reads. (Du et al., 2010)

```{r, echo = FALSE}
# Compute beta values
beta_values <- getBeta(gmSet)
colnames(beta_values) <- sample_sheet$`Sample Name`
```

M values are log transformed beta values. Negative M values indicate higher levels of unmethylated DNA copies than methylated DNA copies. Postive M values indicate higher levels of methylated DNA copies than unmethylated. M values close to zero indicate similar intensities between methylated and unmethylated probes. M values are mathematically expressed as, 

M(i) = log2([max(yi(methyl))+alpha]/[max(yi(unmethyl))+alpha])

where alpha is a constant to offset dramatic changes resulting from small estimation errors. M values can be expressed in terms of beta values,

M(i) = log2(Beta(i)/[1 - Beta(i)])

and therefore, beta values can be expressed in terms of M values,

Beta(i) = 2^M(i) / [2^M(i) + 1]

M values are more statistically sound than beta values.

```{r, echo = FALSE}
Mval <- getM(gmSet)
dim(Mval)
```

It appears that the group with Sentrix ID 204081560070 clusters, suggesting a batch effect?
```{r, echo = FALSE}
# PCA plot of M values
par(mfrow=c(1,1))
plotMDS(Mval, labels=sample_sheet$`Sample Name`, col=as.integer(factor(sample_sheet$Status)))
legend("top",legend=c("Control","Patient"),pch=16,cex=.5,col=1:2)
```

## Differential Methylation Analysis

Design matrix

```{r, echo = FALSE}
batch <- factor(ifelse(sample_sheet$`Sentrix ID` == 204081560070, "batch.effect", "no.batch"))
disease <- factor(sample_sheet$Status)
design <- model.matrix(~1 + disease + batch, data = sample_sheet)
colnames(design) <- c(levels(disease),levels(batch)[-1])
design
```

Fit model using M values as M values are more statistically valid, beta is more intuitive (Du et al., 2010)
```{r}
fit.reduced <- lmFit(Mval, design)
# Contrast matrix
colnames(design) <- make.names(colnames(design))
contrast <- makeContrasts(control - patient, levels = design)
contrast
fit <- contrasts.fit(fit.reduced, contrast)
fit <- eBayes(fit)
```
Model Results
```{r, echo = FALSE}
summary(decideTests(fit))
top<-topTable(fit,coef=1, number = Inf)
head(top)
```

Visualize results using beta values

```{r, echo = FALSE}
cpgs <- rownames(top)
par(mfrow = c(2,2), cex = 0.5)
for(i in 1:4){
stripchart(beta_values[rownames(beta_values) %in% cpgs[i],]~design[,2],method="jitter",
group.names=c("Control","Patient"), pch = 16,cex=1.5,col=c(4,2), ylab="Beta values",
vertical=TRUE,cex.axis=1.5,cex.lab=1.5)
title(cpgs[i],cex.main=1.5)
}
```

## Differential Variability Analysis

fit model testing variances between controls and patients

```{r, echo = FALSE}
fitvar <- varFit(Mval, design = design, coef = 1:3)
summary(decideTests(fitvar))
topVar <- topVar(fitvar, coef=3)
topVar
```

Visualize top 4 most differentially variable CpGs

```{r, echo = FALSE}
cpgsDV <- rownames(topVar)
par(mfrow=c(2,2), cex = 0.5)
for(i in 1:4){
stripchart(beta_values[rownames(beta_values)==cpgsDV[i],]~design[,2], method="jitter",
group.names=c("Control","Patient"),pch=16,cex=1.5,col=c(4,2),ylab="Beta values",
vertical=TRUE,cex.axis=1.5,cex.lab=1.5)
title(cpgsDV[i],cex.main=1.5)
}
```

Identify the most variable patient sample

```{r, echo = FALSE}
par(mfrow=c(2,2), cex = 0.5)
for(i in 1:4){
stripchart(beta_values[rownames(beta_values)==cpgsDV[i],]~colnames(beta_values), method="jitter",pch=16,cex=1.5,ylab="Beta values",
vertical=TRUE,cex.axis=1.5,cex.lab=1.5)
title(cpgsDV[i],cex.main=1.5)
}
```

## Gene Ontology Analysis

### CpG Analysis

```{r, echo = F}
# Observe differentially methylated CpGs using 1% FDR cut-off
table(top$adj.P.Val < 0.01)
```

505k significant cpg sites is too much to yield biologically relevant results. We can trim the data by introducing a cutoff of the change in beta values.
```{r}
# make sure that order of beta values matches order after analysis
beta <- getBeta(gmSet)
beta <- beta[match(rownames(top),rownames(beta)),]
beta_ctrl <- rowMeans(beta[,disease=="control"])
beta_pt <- rowMeans(beta[,disease=="patient"])
Delta_beta <- beta_ctrl - beta_pt
sigDM <- top$adj.P.Val < 0.01 & abs(Delta_beta) > 0.25
table(sigDM)
```


```{r}
topCpGs <- beta[sigDM == TRUE,]
sigCpGs <- rownames(topCpGs)
check <- getMappedEntrezIDs(sig.cpg = sigCpGs, array.type = "EPIC")
```

```{r}
# Check number of genes to which CpGs are annotated
length(check$sig.eg)
```

```{r}
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
gst <- gometh(sig.cpg=sigCpGs, all.cpg=rownames(top), collection="GO", array.type = "EPIC",
              plot.bias=TRUE) # Include "all.cpg" option if filtering CpGs prior to testing
```

```{r}
topGSA(gst)
```

We can also test according to a specified gene set using "gsameth"

### Region level analysis

```{r}
source("/Users/kennedyzinn/Desktop/OneDrive - UTHealth Houston/GRA/dna methyl/region_analysis.R")
region_analysis("control - patient", "GO", NULL, "0.01")
```

We can also test according to a specified gene set using "gsaregion"

## KEGG Analysis

### CpG Analysis

```{r}
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
gst <- gometh(sig.cpg=sigCpGs, all.cpg=rownames(top), collection="KEGG", array.type = "EPIC",
              plot.bias=TRUE) # Include "all.cpg" option if filtering CpGs prior to testing
```

```{r}
topGSA(gst)
# We can also test according to a specified gene set using "gsameth"
```


### Region level analysis

```{r}
region_analysis("control - patient", "KEGG", NULL, "0.01")
```

```{r}
# plot DMRs
cols <- c(1,2)[disease]
names(cols) <-disease
par(mfrow=c(1,1))
DMR.plot(ranges=results.ranges, dmr=2, CpGs=beta_values, phen.col=cols, 
           what="Beta", arraytype="EPICv1", genome="hg19")
```

